Shallow EDSLs and Object-Oriented Programming

### 2016-06-22 paper/sections/Introduction.tex
Intro

Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in OOP. Would this OO approach be
classified as a shallow or deep embedding? We believe arguments can be
made both ways. Since the {\sc Composite} or {\sc Interpreter}
patterns are normally accepted to provide a way to encode ASTs, it would be
reasonable to say that \emph{according to definition of deep embedding
  above, the OO approach classifies as a deep
  embedding}.

@@

In this paper, however, we prefer to interpret shallow embeddings
differently. We argue that a straightforward OO approach is really a
shallow embedding. The reason is that shallow embeddings are typically
implemented using procedural abstraction and, as Cook~\cite{} argued,
procedural abstraction is the essence of Object-Oriented
Programming. Thus the implementation of a shallow
DSL in OOP languages should simply correspond to a standard
object-oriented program.

If we accept this view, 
a natural way to distinguish implementations of EDSLs is 
in terms of the data abstraction used to model the language
constructs instead. We can follow Reynold's~\cite{} classification of
data abstractions: \emph{procedural abstraction} and \emph{user-defined
  types}. It is clear that shallow embeddings use \emph{procedural
  abstraction}~\cite{}: the DSLs are modelled by interpretation
functions. Thus, the other implementation option for EDSLs is to
use \emph{user-defined types}. In Reynolds terminology user-defined
types mean disjoint union types, which are nowadays commonly available
in modern languages as \emph{algebraic datatypes}. Disjoint union
types can also be emulated in OOP using the {\sc Visitor} pattern.

A distinction based on data abstraction is more precise and provides a
remedy for possible misinterpretation. An EDSL implemented with
algebraic datatypes falls into the category of user-defined types
(deep embedding), while a Composite-based OO implementation falls
under procedural abstraction (shallow embedding).

@@

As Cook~\cite{} argued
procedural abstraction is the essence of Object-Oriented
Programming. If we accept Cook's view, the implementation of a shallow
DSL in OOP languages should simply correspond to a standard
object-oriented program. Given that OOP languages have
evolved over more than 50 years to improve the use of procedural
abstraction, they ought to have some advantages to encode shallow
EDSLs.

The main goal of this paper is to show that OOP languages have advantages for the
implementation of shallow embeddings.  
In order to understand how OOP can help with shallow embeddings, lets
first review some of the limitations of shallow embeddings commonly found in
the literature:

@@

We show that OOP abstractions, including \emph{(multiple)
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{}, which provides a simple
solution to the \emph{Expression Problem}\cite{} in OOP languages. Thus
using just standard OOP mechanims enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to Haskell.

We make our arguments by taking a recent paper by Gibbons and Wu,
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala. Although from the
\emph{syntactical} point of view there are minor inconveniences in
the Scala version, from the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages.

In summary, our contributions are:

@@

\item {Identifying Advantages of OOP languages for Shallow Embeddings:}

@@

\end{itemize}

>>>>

Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in OOP. Would this OO approach be
classified as a shallow or deep embedding? We believe arguments can be
made both ways. The {\sc Composite} or {\sc Interpreter}
patterns are normally accepted to provide a way to encode ASTs. Thus, 
one possible interpretation is that \emph{according to definition of deep embedding
  above, the OO approach classifies as a deep
  embedding}.

@@

In this paper, however, we prefer to interpret shallow and deep embeddings
differently. We argue that a straightforward OO approach is really a
shallow embedding. The reason is that shallow embeddings are 
implemented using \emph{procedural abstraction} and, as Cook~\cite{cook09abstraction} argued,
procedural abstraction is the essence of Object-Oriented
Programming. Thus the implementation of a shallow
DSL in OOP languages should simply correspond to a standard
object-oriented program.

If we accept Cook's view on OOP, 
a natural way to distinguish implementations of EDSLs is 
in terms of the data abstraction used to model the language
constructs instead. As Reynold's~\cite{reynolds94proceduralabstraction} suggested there are two
types of data abstraction: procedural abstraction and \emph{user-defined
  types}. It is clear that shallow embeddings use \emph{procedural
  abstraction}: the DSLs are modelled by interpretation
functions. Thus, the other implementation option for EDSLs is to
use \emph{user-defined types}. In Reynolds terminology user-defined
types mean disjoint union types, which are nowadays commonly available
in modern languages as \emph{algebraic datatypes}. Disjoint union
types can also be emulated in OOP using the {\sc Visitor} pattern.

A distinction based on data abstraction is more precise and provides a
remedy for possible misinterpretation. An EDSL implemented with
algebraic datatypes falls into the category of user-defined types
(deep embedding), while a {\sc Composite}-based OO implementation falls
under procedural abstraction (shallow embedding).

@@

The main goal of this paper is to argue that OOP languages have
advantages for the implementation of shallow embeddings. This should
not came as a surprise, as OOP languages have evolved over more than 50
years to improve the use of procedural abstraction.

In order to understand how OOP can help with shallow embeddings, lets
first review some of the limitations of shallow embeddings commonly
found in the literature:

@@

We show that OOP abstractions, including \emph{(multiple)
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanims enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to
Haskell.
 \bruno{double-check if the paper really shows this!}

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala. Although from the
\emph{syntactical} point of view there are minor inconveniences in
the Scala version, from the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages.

\begin{comment}
In summary, our contributions are:

@@

\item {}

@@

\end{itemize}
\end{comment}
